---
title: "Valuing dung beetles on Florida cattle ranches"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Valuing dung beetles on Florida cattle ranches}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(RancheR)
```

## What this package does

`RancheR` estimates the **annual economic benefit that dung beetles provide to a
cattle operation**. Dung beetles break down and bury dung pats; without them,
pats sit on the pasture for far longer, fouling grass that cattle then refuse to
graze. By accelerating pat decay, beetles free up grazing area — effectively
raising the carrying capacity of a ranch.

The package turns that ecological service into dollars, using empirical
pat-decay data from Central Florida pasture research (Stanbrook-Buyer, Bhat &
King, 2024). It contains the same calculations that power the
[RancheR Shiny app](https://rstanbrook-rancher.hf.space/), exposed as plain R
functions so you can script them, batch them, or build your own tools on top.

## The model in brief

The calculation follows four steps:

1. **Decay rate.** Each beetle-abundance *scenario* has an empirical pat decay
   rate (grams/day). More beetles → faster decay.
2. **Fouled pasture.** Faster decay means fewer grazing-area units (GAU) of
   pasture are fouled per cow per year. The package stores the empirically
   measured fouled area for each scenario.
3. **Avoided fouling.** Comparing a scenario against the *no-beetle* baseline
   gives the GAU of fouling **avoided** thanks to beetles.
4. **Dollar value.** Avoided fouling is converted to an equivalent number of
   additional cows the freed pasture can support, then multiplied by ranch-level
   income per cow.

## A first estimate

The main function is `calc_dung_beetle_benefit()`. The only argument you must
supply is the herd size; everything else has a sensible paper-derived default.

```{r}
res <- calc_dung_beetle_benefit(
  num_cattle = 100,
  scenario   = "Managed (low beetle abundance)"
)

res
```

The return value is a list. The two headline numbers are usually:

```{r}
res$annual_value     # estimated annual benefit, in dollars
res$additional_cows  # equivalent extra cows the freed pasture supports
```

## The three scenarios

There are three beetle-abundance scenarios. You can list them with
`dung_scenarios()`:

```{r}
dung_scenarios()
```

Their decay rates (grams of dung removed per day) are:

```{r}
sapply(dung_scenarios(), scenario_decay_rate)
```

Because more beetles remove dung faster, the economic benefit rises from the
"no beetles" baseline up to the "natural" high-abundance case. Here is the
benefit for a 100-head operation under each scenario:

```{r}
herd <- 100
benefits <- sapply(dung_scenarios(), function(s) {
  calc_dung_beetle_benefit(num_cattle = herd, scenario = s)$annual_value
})

data.frame(
  scenario     = dung_scenarios(),
  annual_value = round(benefits, 2),
  row.names    = NULL
)
```

The "no beetles" scenario is the baseline, so its avoided fouling — and
therefore its benefit — is zero by construction.

## Tailoring the estimate to your ranch

Every parameter can be overridden. For example, a larger operation with a
higher income per cow and slightly more pasture required per head:

```{r}
calc_dung_beetle_benefit(
  num_cattle     = 500,
  scenario       = "Natural (high beetle abundance)",
  income_per_cow = 900,
  acre_per_cow   = 2.6
)$annual_value
```

The benefit scales linearly with herd size, so doubling the cattle doubles the
estimate:

```{r}
a <- calc_dung_beetle_benefit(250, "Natural (high beetle abundance)")$annual_value
b <- calc_dung_beetle_benefit(500, "Natural (high beetle abundance)")$annual_value
c(herd_250 = a, herd_500 = b, ratio = b / a)
```

### Custom decay rates

If you have your own measured decay rate, pass it via `decay_override`. When you
do, the fouled-area is scaled from the no-beetle baseline by relative decay time
rather than read from the empirical scenario table — letting you explore
conditions between or beyond the measured scenarios (e.g. warm, wet weather
speeds decay; cool, dry weather slows it).

```{r}
calc_dung_beetle_benefit(
  num_cattle     = 100,
  scenario       = "Managed (low beetle abundance)",
  decay_override = 9.5
)$annual_value
```

## Inspecting the underlying constants

All paper-derived defaults and empirical values live in one documented object,
`rancher_constants`, so nothing is hidden inside the functions:

```{r}
rancher_constants
```

## Batch analysis example

Because the functions are plain R, scaling up to many ranches is
straightforward. Here we estimate the benefit across a range of herd sizes under
the natural scenario:

```{r}
herds <- c(50, 100, 250, 500, 1000)
sapply(herds, function(n) {
  round(calc_dung_beetle_benefit(n, "Natural (high beetle abundance)")$annual_value, 2)
})
```

## Adjusting decay rates for your own climate

The decay rates above were measured at a single Central Florida site
(`r rancher_constants$ref_lat`, `r rancher_constants$ref_lon`) during summer,
when temperature and rainfall strongly drive dung decomposition. Decay is
slower in cooler or drier regions and faster in hotter, wetter ones, so a ranch
elsewhere needs locally adjusted rates.

`estimate_local_decay()` does this by scaling the reference rates by how a
location's climate differs from the Florida site. Climate comes from two
[WorldClim](https://www.worldclim.org/) "bioclim" variables chosen to match the
summer collection window:

- **BIO10** — mean temperature of the warmest quarter (&deg;C)
- **BIO18** — precipitation of the warmest quarter (mm)

Temperature acts through a **Q10 response** (decay changes by a factor `q10`,
default 2, per 10 &deg;C — a standard description of biological decomposition),
and moisture through a **saturating response** (decay rises with rainfall but
with diminishing returns). Both are normalised to 1 at the reference climate, so
the model reproduces the measured rates exactly at the Florida site.

> **A note on the method.** Because decay was measured at only one location,
> this is a *calibrated, literature-grounded* scaling — not a regression fitted
> to multi-site data. The `q10` and `precip_half_sat` arguments are exposed so
> the relationship can be tuned or replaced as local field data become
> available.

### Looking up climate automatically

Given a latitude and longitude, `estimate_local_decay()` downloads (and caches)
the WorldClim layers and returns climate-adjusted decay rates for all three
scenarios. This requires the `terra` package and an internet connection on first
use:

```{r, eval = FALSE}
est <- estimate_local_decay(lat = 31.5, lon = -97.1)  # central Texas
est$decay_rates
```

### Supplying climate directly

If you already know the warmest-quarter temperature and precipitation (or want
to work offline), pass them in via `climate` — no download needed:

```{r}
est <- estimate_local_decay(
  lat = 44.0, lon = -89.5,                                   # central Wisconsin
  climate = c(temp_warmq_C = 20.3, precip_warmq_mm = 290)
)

est$climate          # climate used
est$climate_factor   # combined scaling vs. the Florida reference
round(est$decay_rates, 2)
```

The cooler, drier Wisconsin climate scales decay below the Florida baseline.

### Feeding the climate into the economic model

Pass the location's `climate_factor` into `calc_dung_beetle_benefit()`. This
composes correctly with the economic model: slower decay (a factor below 1)
fouls pasture for proportionally longer under *both* the no-beetle baseline and
the beetle scenario, so the avoided fouling — and the dollar benefit — scales by
`1 / climate_factor`. Where decay is slower, beetles avert more fouled-pasture
time, so the modelled benefit is *higher* than at the warmer Florida reference.

```{r}
calc_dung_beetle_benefit(
  num_cattle     = 200,
  scenario       = "Managed (low beetle abundance)",
  climate_factor = est$climate_factor
)$annual_value
```

(Note: `decay_override` is for manual single-rate exploration and is measured
against the *fixed* Florida no-beetle baseline, so it does not compose with the
climate model — use `climate_factor` for location-adjusted economics.)

You can also inspect the raw bioclim values for a point with
`bioclim_warmest_quarter(lat, lon)`.

### Visualising the climate response

The scaling factor is easiest to read as a surface over the two bioclim
variables. The contours below show how decay speeds up toward the warm, wet
corner and slows toward the cool, dry corner; the white point marks the Florida
reference, where the factor is exactly 1.

```{r climate-surface, fig.width = 7, fig.height = 5, fig.alt = "Contour plot of the dung decay climate scaling factor over warmest-quarter temperature and precipitation."}
temps <- seq(15, 35, length.out = 60)   # warmest-quarter temperature (deg C)
precs <- seq(50, 900, length.out = 60)  # warmest-quarter precipitation (mm)
grid  <- outer(temps, precs, function(t, p) climate_decay_factor(t, p)$climate_factor)

filled.contour(
  temps, precs, grid,
  color.palette = function(n) hcl.colors(n, "YlOrRd", rev = TRUE),
  xlab = "Warmest-quarter temperature (°C)",
  ylab = "Warmest-quarter precipitation (mm)",
  main = "Climate scaling of dung decay (relative to Florida reference)",
  plot.axes = {
    axis(1); axis(2)
    points(rancher_constants$ref_temp_warmq_C,
           rancher_constants$ref_precip_warmq_mm,
           pch = 21, bg = "white", cex = 1.6)
    text(rancher_constants$ref_temp_warmq_C,
         rancher_constants$ref_precip_warmq_mm,
         "FL reference", pos = 4, cex = 0.8)
  }
)
```

### Mapping a few locations

Applied geographically, the same model gives a decay rate for any ranch. The map
below places a handful of sites and labels each with its estimated
managed-scenario rate (g/day). The warmest-quarter climate values here are
**illustrative approximations** for demonstration — in practice
`estimate_local_decay(lat, lon)` looks up the real WorldClim values for you.

```{r decay-map, eval = requireNamespace("maps", quietly = TRUE), fig.width = 7, fig.height = 4.6, fig.alt = "Map of the contiguous United States with example sites coloured by estimated managed-scenario dung decay rate."}
sites <- data.frame(
  name            = c("FL reference", "S. Texas", "Georgia", "Wisconsin", "Colorado"),
  lon             = c(-81.20, -98.49, -83.40, -89.50, -105.00),
  lat             = c( 27.13,  27.80,  32.80,  44.00,   40.30),
  temp_warmq_C    = c( 27.31,  29.50,  26.80,  20.30,   17.50),
  precip_warmq_mm = c(   522,    210,    330,    290,     140)
)

sites$decay_managed <- vapply(seq_len(nrow(sites)), function(i) {
  estimate_local_decay(
    lat = sites$lat[i], lon = sites$lon[i],
    climate = c(temp_warmq_C = sites$temp_warmq_C[i],
                precip_warmq_mm = sites$precip_warmq_mm[i])
  )$decay_rates[["Managed (low beetle abundance)"]]
}, numeric(1))

pal  <- hcl.colors(100, "YlOrRd", rev = TRUE)
cols <- pal[cut(sites$decay_managed, breaks = 100, labels = FALSE)]

maps::map("state", fill = TRUE, col = "grey95", border = "grey80")
points(sites$lon, sites$lat, pch = 21, bg = cols, cex = 2.4)
text(sites$lon, sites$lat, sprintf("%.1f", sites$decay_managed),
     pos = 3, cex = 0.75)
title("Estimated managed-scenario decay rate (g/day)")
```

## Citation

If you use `RancheR` in your work, please cite:

> Stanbrook-Buyer, R., Bhat, M., & King, J. R. (2024). Economic value of dung
> removal by dung beetles in US sub-tropical pastures. *Basic and Applied
> Ecology*, *79*, 123–130. https://doi.org/10.1016/j.baae.2024.07.001
